---
title: "ggplot-jamboree-heart"
author: "Alison Hill"
output:
  html_document:
    keep_md: TRUE
    highlight: pygments
    theme: flatly
    smart: false
    toc: TRUE
    toc_float: TRUE
---

```{r}
library(readr)
library(ggplot2)
library(ggbeeswarm)
```

Read in the csv from link
```{r}
heart <- read_csv("http://faculty.washington.edu/kenrice/heartgraphs/nhaneslarge.csv", na = ".")
```


# Univariate Plots: Folate intake by gender

```{r}
ggplot(heart, aes(x = DR1TFOLA)) +
  geom_histogram()
```

Update x-axis title
```{r}
ggplot(heart, aes(x = DR1TFOLA)) +
  geom_histogram() +
  labs(x = "Folate intake")
```

Change outline of bars
```{r}
ggplot(heart, aes(x = DR1TFOLA)) +
  geom_histogram(colour = "white") +
  labs(x = "Folate intake")
```

Change fill of bars
```{r}
ggplot(heart, aes(x = DR1TFOLA)) +
  geom_histogram(colour = "white", fill = "peachpuff") +
  labs(x = "Folate intake")
```

Change number of breaks to 50
```{r}
ggplot(heart, aes(x = DR1TFOLA)) +
  geom_histogram(colour = "white", fill = "peachpuff", bins = 50) +
  labs(x = "Folate intake")
```

Facet by gender
```{r}
ggplot(heart, aes(x = DR1TFOLA)) +
  geom_histogram(colour = "white", fill = "peachpuff", bins = 50) +
  labs(x = "Folate intake") +
  facet_wrap(~gender)
```

Kernel density instead
```{r}
ggplot(heart, aes(x = DR1TFOLA)) +
  geom_density(colour = "white", fill = "peachpuff") +
  labs(x = "Folate intake") +
  facet_wrap(~gender)
```

Kernel density without facet

Notice moving the colour into aesthetics
```{r}
ggplot(heart, aes(x = DR1TFOLA)) +
  geom_density(aes(colour = gender)) +
  labs(x = "Folate intake") 
```

This is the exact same!
```{r}
ggplot(heart, aes(x = DR1TFOLA, colour = gender)) +
  geom_density() +
  labs(x = "Folate intake") 
```

# Univariate Plots: Systolic blood pressure by gender

Start with simple stripchart of `BPXSAR`

```{r}
ggplot(heart, aes(x = gender, y = BPXSAR)) +
  geom_point() 
```

Too many points! Add alpha
```{r}
ggplot(heart, aes(x = gender, y = BPXSAR)) +
  geom_point(alpha = .1)
```

Try jittering
```{r}
ggplot(heart, aes(x = gender, y = BPXSAR)) +
  geom_jitter(alpha = .1)
```

Change width/height of jitter
```{r}
ggplot(heart, aes(x = gender, y = BPXSAR)) +
  geom_jitter(alpha = .1, width = .5, height = 0)
```

Change x-axis title
```{r}
ggplot(heart, aes(x = gender, y = BPXSAR)) +
  geom_jitter(alpha = .1, width = .5, height = 0) +
  labs(x = "Systolic BP (mmHg)", y = "")
```


Beeswarm-like in ggplot
```{r}
ggplot(heart, aes(x = gender, y = BPXSAR)) +
  geom_beeswarm()+
  labs(x = "Systolic BP (mmHg)", y = "")
```

Add alpha again
```{r}
ggplot(heart, aes(x = gender, y = BPXSAR)) +
  geom_beeswarm(alpha = .2) +
  labs(x = "Systolic BP (mmHg)", y = "")
```

Add the mean plus 95% CI
```{r}
ggplot(heart, aes(x = gender, y = BPXSAR)) +
  geom_beeswarm(alpha = .2) +
  stat_summary(fun.y = "mean", geom = "point", colour = "orange") +
  stat_summary(fun.data = mean_cl_boot, geom = "linerange", colour = "orange") +
  labs(x = "Systolic BP (mmHg)", y = "")
```

Box + beeswarm plot (try re-ordering)
```{r}
ggplot(heart, aes(x = gender, y = BPXSAR)) +
  geom_boxplot(outlier.shape = NA) +
  geom_beeswarm(alpha = .2) +  
  labs(x = "Systolic BP (mmHg)", y = "")
```


Violin plot

```{r}
ggplot(heart, aes(x = gender, y = BPXSAR)) +
  geom_violin(alpha = .2) +
  labs(x = "Systolic BP (mmHg)", y = "")
```

Add sample mean and median
```{r}
ggplot(heart, aes(x = gender, y = BPXSAR)) +
  geom_violin(alpha = .2) +
  stat_summary(fun.y = "mean", geom = "point", colour = "orange") +
  stat_summary(fun.y = "median", geom = "point", colour = "blue") +
  labs(x = "Systolic BP (mmHg)", y = "")
```

Add boxplot
```{r}
ggplot(heart, aes(x = gender, y = BPXSAR)) +
  geom_violin(alpha = .2) +
  geom_boxplot(width = .05) +
  labs(x = "Systolic BP (mmHg)", y = "")
```


# Bivariate Plots: Age & Systolic Blood Pressure

Simple scatterplot
```{r}
ggplot(heart, aes(x = RIDAGEYR, y = BPXSAR)) +
  geom_point() +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)")
```

If you have big n, try hexbin plot
```{r}
ggplot(heart, aes(x = RIDAGEYR, y = BPXSAR)) +
  geom_hex() +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)")
```

Make colors make more sense
```{r}
library(viridis)
ggplot(heart, aes(x = RIDAGEYR, y = BPXSAR)) +
  geom_hex() +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)") +
  scale_fill_gradientn(colours = viridis(256, begin = 1, end = 0))
```


```{r}
library(ggalt)
ggplot(heart, aes(x = RIDAGEYR, y = BPXSAR)) +
  stat_bkde2d(aes(fill = ..level.., alpha = ..level..), geom = "polygon", bandwidth = c(2,2))  +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)") +
  scale_fill_gradientn(colours = viridis(256, begin = 1, end = 0))
```

Add linear regression line with SE
```{r}
ggplot(heart, aes(x = RIDAGEYR, y = BPXSAR)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)")
```

Default is loess line
```{r}
ggplot(heart, aes(x = RIDAGEYR, y = BPXSAR)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)")
```

Add splines
```{r}
library(splines)
library(MASS)
ggplot(heart, aes(x = RIDAGEYR, y = BPXSAR)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ ns(x, 3)) +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)")
```

# Multivariable Plots: Body Mass Index & Systolic BP, by gender and age

Just copy this:
```{r}
library(dplyr)
heart2 <- heart %>% 
  mutate(age_cat = cut(RIDAGEYR,c(0,30,55,100)))
```

Recreate theirs first

```{r}
ggplot(heart2, aes(x = BMXBMI, y = BPXSAR)) +
  geom_point() +
  stat_smooth(aes(colour = gender), method = "lm") +
  facet_wrap(~age_cat) +
  labs(x = "Body Mass Index"~(kg/m^2), y = "Systolic BP (mmHg)")
```

Try with facet grid, update labels

```{r}
ggplot(heart2, aes(x = BMXBMI, y = BPXSAR)) +
  geom_point() +
  stat_smooth(aes(colour = gender), method = "lm") +
  facet_grid(gender~age_cat) +
  labs(x = "Body Mass Index"~(kg/m^2), y = "Systolic BP (mmHg)")
```

Play with colors!
```{r}
ggplot(heart2, aes(x = BMXBMI, y = BPXSAR, colour = gender)) +
  geom_point(alpha = .5) +
  stat_smooth(method = "lm") +
  facet_grid(gender~age_cat) +
  theme_minimal() +
  labs(x = "Body Mass Index"~(kg/m^2), y = "Systolic BP (mmHg)") +
  scale_color_manual(values = c("#B47CC7", "#D65F5F"), guide = FALSE)
```


```{r}
my_colors <- c("#C4AD66", "#77BEDB")
ggplot(heart2, aes(x = BMXBMI, y = BPXSAR, colour = gender)) +
  geom_point(alpha = .5) +
  stat_smooth(method = "lm") +
  facet_grid(gender~age_cat) +
  labs(x = "Body Mass Index"~(kg/m^2), y = "Systolic BP (mmHg)") +
  scale_color_manual(values = my_colors, guide = FALSE)
```


